A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland

### 2022-02-03 paper3.Rmd
Bourgeois-Pichat is back

# Neonatal mortality

@@

The analysis is split into two parts by season, *winter + spring* is one, and *summer + fall* the other.

## Winter and Spring

@@

## Summer and Fall

@@

# Postneonatal mortality

@@

## Bourgeois-Pichat and the log-cube transform

**FILL IN ...**

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                    lowTemp + highTemp + season + aver
                + emeantemp + subreg, 
                data = pneo, cuts = g(c(28, 365)))
```

## Winter and Spring

```{r postanalw, results='asis'}
winter <- tpneo[tpneo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)

@@

## Summer and Fall

```{r postanals, results='asis'}
summer <- tpneo[tpneo$season %in% c("summer", "fall"), ]
winter$season <- factor(winter$season)

@@

>>>>

# Results

The results regarding neonatal mortality is much in accordance with the results
found by @junkka. However, they used temperature in a "hockey-stick" regression with a breakpoint at
14.5 degrees Celsius and a negative slope (decreasing risk) to the left and a positive slope
(increasing risk) to the right. Instead, we are using the average weekly temperature
for the 52 weeks of a year, for each week averaging over all the years in the study, as
our "reference points" ("climate"), adding deviances up and down ("weather") as
"short-term temperature stress". This is similar to the way prices and mortality were related
in for instance @genus11, that is, a time series split into long time trend and short
term variation.

In accordance with the results of @bp51a and @bp51b, we separate the investigation into 
two parts, *neonatal* and *postneonatal* mortality.

## Neonatal mortality

@@

The analyses are split into two parts by season, *winter + spring* is one, and *summer + fall* the other.

### Winter and Spring

@@

### Summer and Fall

@@

## Postneonatal mortality

@@

### Bourgeois-Pichat and the log-cube transform

The *log-cube transform g*, suggested by @bp51a, is defined as

\begin{equation}
g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
\end{equation}

where $t$ is age in days. A graph of the transform is shown in Figure \@ref(fig:gplot).

```{r gplot, fig.cap = "The g transform.",fig.height=3.2}
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", ylab = "g(t)", xlab = "t (days)", col = "blue", axes = FALSE)
abline(h = 0, v = 0)
##
lines(c(28, 28), c(0, g(28)), lty = 3)
lines(c(0, 28), c(g(28), g(28)), lty = 3)
##
lines(c(183, 183), c(0, g(183)), lty = 3)
lines(c(0, 183), c(g(183), g(183)), lty = 3)
##
lines(c(365, 365), c(0, g(365)), lty = 3)
lines(c(0, 365), c(g(365), g(365)), lty = 3)
##
axis(1, at = c(0, 28, 183, 365))
ycor <-  c(0, g(28), g(183), g(365))
axis(2, at = ycor, labels = round(ycor), las = 1)
box()
```

It was used as a tool for dividing infant mortality 
into *endogenous* and *exogenous* causes, and a an important part in that venue was
the observations (i) postneonatal mortality is purely exogenous, and (ii) the distribution 
of exogenous mortality is *truncated uniform* (right truncated at $t = g(365)$) on the
$g$-transformed time scale.

Today, assumption (ii) is slightly outdated, for three reasons: (a) The uniform distribution
is not a very practical tool in survival analysis, (b) replacing the uniform with an
*exponential* not only fits better (usually), it also is a survival distribution that
is extremely easy to work with in survival analysis, and (c) "all models are wrong, but some are useful [@box76]".

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
gpneo <- pneo
gpneo$enter <- g(gpneo$enter * 365)
gpneo$exit <- g(gpneo$exit * 365)
##
gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                    lowTemp + highTemp + season + aver
                + emeantemp + subreg, 
                data = gpneo, cuts = g(c(28, 365)))
```

For now, let us look at the basic distribution of postneonatal life on the $g$ time
scale. We do it nonparametrically by calculating the *Nelson-Aalen* estimator of the
cumulative hazards function. See Figure \@ref(fig:nap), where the two curves are indistinguishable.

```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
plot(fit, xlam = "g(t)")
fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
zz <- hazards(fit.w)
lines(zz$x, zz$y, col = "red", lty = 2)
legend("topleft", legend = c("Nelson-Aalen", "Exponential"), lty = 1:2, col = c("blue", "red"))
```

### Winter and Spring

```{r postanalw, results='asis'}
winter <- gtpneo[gtpneo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)

@@

### Summer and Fall

```{r postanals, results='asis'}
summer <- gtpneo[gtpneo$season %in% c("summer", "fall"), ]
winter$season <- factor(winter$season)

@@

# The biometric analysis of infant mortality

A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
*without* the "log-cube transform".
